---
title: "Workflow to produce visualizations of Pozniak, et al 2022 mouse melanoma data"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This workflow displays a variety of interactive Plotly visualizations to explore
the *mouse* melanoma data published in the recent [Pozniak, et al 2022](https://www.biorxiv.org/content/10.1101/2022.08.11.502598v1).
Plotly visualizations found in this workflow are also found in the accompanying 
R Shiny App. 

```{r message=FALSE, warning=FALSE}
library("dplyr")
library("tibble")
library("ggplot2")
library("ggpubr")
library("GSVA")
library("readr")
library("heatmaply")
library("plotly")
```

### Load data

The aggregated `NRAS13_all_43k` cell dataset and test gene set consisting of two
gene files are loaded:

```{r message=FALSE, warning=FALSE}
## read aggregated count human malignant dataset 
df1 <- readRDS(file = "./counts/NRAS13_all_43k_agg_counts.rds")
# df2 <- readRDS(file = "./counts/NRAS13_malign_preprint_agg_counts.rds")
# head(df1)
```

### Gene lists

This workflow is primarily designed with the intention to correlate two gene
sets. Here we load such a data file with two toy gene sets. It is possible to 
upload a custom user-generated gene set as long as the file is a CSV or TSV file
and the gene sets are the first and second columns. This workflow ignores
additional columns, and these columns may be named whatever the end user wishes.

```{r message=FALSE, warning=FALSE}
## read gene sets of interest to assay in the dataset
g1 <- readr::read_csv("./counts/mouse_test_genes.csv")

## display gene lists as data table
g1 %>% 
  DT::datatable()
```

### Create gene list

This block of code puts each row of the test gene set into a list with each
element in the list given the same name as the original dataframe columns,
however `NA` values are removed from the list elements.

```{r message=FALSE, warning=FALSE}
gl <- list()
for (i in 1:ncol(g1)) {
  g1 %>% 
    dplyr::select(i) %>%
    dplyr::pull() -> gl[[i]]
  gl[[i]] <- gl[[i]][!is.na(gl[[i]])]
}
names(gl) <- names(g1)
```

### Run GSVA

The app applies GSVA on the gene sets found in the test gene file. This is the
most time consuming computational step in the app.

```{r message=FALSE, warning=FALSE}
## run ssGSEA
GSVA::gsva(expr = as.matrix(df1 %>% dplyr::select(-gene) %>% as.data.frame()),
           gset.idx.list = gl) -> gsva_sig
```

## Violin Plots

The dataframe resulting from the GSVA step is converted to a long dataframe and
the information contained in the aggregated cell names is used to re-label the
aggregated cell types by sample.

```{r message=FALSE, warning=FALSE}
## convert data for violin Plot 
gsva_sig %>%
  as.data.frame() %>%
  tibble::rownames_to_column() %>%
  dplyr::rename(gene_sig = names(.)[1]) %>%
  tidyr::gather(key = "cell_name", value = "expression", -gene_sig) %>%
  dplyr::mutate(sample = gsub("^[^_]*.", "",
                              gsub("_[^_]+$", "", cell_name))) %>%
  dplyr::mutate(cell_type = dplyr::case_when(
    grepl("B-cell", cell_name) ~ "B-cell",
    grepl("CAF", cell_name) ~ "CAF",
    grepl("DC", cell_name) ~ "DC",
    grepl("EC", cell_name) ~ "EC",
    grepl("Malignant", cell_name) ~ "Malignant",
    grepl("Monocyte", cell_name) ~ "Monocyte/macrophage",
    grepl("Pericyte", cell_name) ~ "Pericyte",
    grepl("NKcell", cell_name) ~ "T/NKcell")) %>%
  as.data.frame() -> a1
```

### Violin plots by gene set score

This plot displays gene expression by cell type for each gene signature. Due to 
UI space constraints in the app, we will probably be limited to displaying only
the first two gene sets.

```{r message=FALSE, warning=FALSE}
a1 %>%
  ggplot(aes(x = cell_type, y = expression, fill = cell_type)) +
  geom_violin(alpha = 0.8) +
  geom_point(position = position_jitter(seed = 1, width = 0.2),
             size = 0.4, alpha = 0.8) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Violin plot of gene signature scores by cell type") +
  facet_grid(~gene_sig) -> p1
ggplotly(p1)
# p1
```

## Heatmaps

This block of code selects all of the genes found in the text file, extracts the
counts for each sample and averages them for these genes. The gene counts have
been transformed to log scale in order for highly expressed genes not skew the 
gradient scale and drown out subtle changes in expression.

```{r message=FALSE, warning=FALSE}
df1 %>%
  dplyr::filter(gene %in% unlist(g1)[!is.na(unlist(g1))]) %>%
  tidyr::gather(key = "cell_name", value = "expression", -gene) %>%
  dplyr::mutate(sample = gsub("^[^_]*.", "",
                              gsub("_[^_]+$", "", cell_name))) %>%
  dplyr::mutate(cell_type = dplyr::case_when(
    grepl("B-cell", cell_name) ~ "B-cell",
    grepl("CAF", cell_name) ~ "CAF",
    grepl("DC", cell_name) ~ "DC",
    grepl("EC", cell_name) ~ "EC",
    grepl("Malignant", cell_name) ~ "Malignant",
    grepl("Monocyte", cell_name) ~ "Monocyte/macrophage",
    grepl("Pericyte", cell_name) ~ "Pericyte",
    grepl("NKcell", cell_name) ~ "T/NKcell")) %>%
  dplyr::group_by(gene, cell_type) %>%
  dplyr::summarise(mean_exp = mean(expression)) %>%
  tidyr::spread(key = "cell_type", value = "mean_exp") %>%
  tibble::column_to_rownames(var = "gene") %>%
  dplyr::mutate_all(log) %>%
  as.data.frame() -> r1
is.na(r1) <- sapply(r1, is.infinite)
r1[is.na(r1)] <- 0

heatmaply::heatmaply(
    x = r1,
    main = "Heatmap of log gene counts",
    xlab = "Cell types",
    ylab = "Genes of interest",
    scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(
      low = "blue",
      high = "firebrick"))
```

### Correlation log gene expression heatmap

```{r message=FALSE, warning=FALSE}
df1 %>%
  dplyr::filter(gene %in% unlist(g1)[!is.na(unlist(g1))]) %>%
  tidyr::gather(key = "cell_name", value = "expression", -gene) %>%
  dplyr::mutate(sample = gsub("^[^_]*.", "",
                              gsub("_[^_]+$", "", cell_name))) %>%
  dplyr::mutate(cell_type = dplyr::case_when(
    grepl("B-cell", cell_name) ~ "B-cell",
    grepl("CAF", cell_name) ~ "CAF",
    grepl("DC", cell_name) ~ "DC",
    grepl("EC", cell_name) ~ "EC",
    grepl("Malignant", cell_name) ~ "Malignant",
    grepl("Monocyte", cell_name) ~ "Monocyte/macrophage",
    grepl("Pericyte", cell_name) ~ "Pericyte",
    grepl("NKcell", cell_name) ~ "T/NKcell")) %>%
  dplyr::group_by(gene, cell_type) %>%
  dplyr::summarise(mean_exp = mean(expression)) %>%
  tidyr::spread(key = "cell_type", value = "mean_exp") %>%
  tibble::column_to_rownames(var = "gene") %>%
  dplyr::mutate_all(log) %>%
  as.data.frame() -> r2
is.na(r2) <- sapply(r2, is.infinite)
r2[is.na(r2)] <- 0

r2 %>%
  t() %>%
  cor() %>%
heatmaply::heatmaply(
    main = "Correlation Heatmap of log gene counts across samples",
    xlab = "Cell types by time point",
    ylab = "Genes of interest",
    scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(
      low = "blue",
      high = "firebrick"))
```

### Correlation heatmap across samples

```{r message=FALSE, warning=FALSE}
r2 %>%
  cor() %>%
heatmaply::heatmaply(
    main = "Correlation Heatmap of log gene counts across samples",
    xlab = "Cell types by time point",
    ylab = "Genes of interest",
    scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(
      low = "blue",
      high = "firebrick"))
```

## Correlation plots

The following correlation plots display how each gene set correlate each gene
set with each other across cell types and time points. The correlation for each
group is reported by displaying the R value and p-value.

```{r message=FALSE, warning=FALSE}
gsva_sig %>%
  t() %>% as.data.frame() %>%
  tibble::rownames_to_column() %>%
  dplyr::rename(sample_name = names(.)[1]) %>%
  dplyr::mutate(sample = gsub("^[^_]*.", "",
                              gsub("_[^_]+$", "", sample_name))) %>%
  dplyr::mutate(cell_type = dplyr::case_when(
    grepl("B-cell", sample_name) ~ "B-cell",
    grepl("CAF", sample_name) ~ "CAF",
    grepl("DC", sample_name) ~ "DC",
    grepl("EC", sample_name) ~ "EC",
    grepl("Malignant", sample_name) ~ "Malignant",
    grepl("Monocyte", sample_name) ~ "Monocyte/macrophage",
    grepl("Pericyte", sample_name) ~ "Pericyte",
    grepl("NKcell", sample_name) ~ "T/NKcell")) %>%
  as.data.frame() -> d1
```

### Correlation between signature scores combined across cell types and timepoints

```{r message=FALSE, warning=FALSE}
d1 %>%
  dplyr::group_by(sample) %>%
  dplyr::summarise(mean_gene1 = mean(gene1),
                   mean_gene2 = mean(gene2)) %>%
  ggplot(aes(x = mean_gene1, y = mean_gene2)) +
  geom_point(size = 1, color = "black") +
  stat_smooth(method = "lm", se = TRUE, fill = "gray", color = "darkgray",
              formula = y ~ poly(x, 1, raw = TRUE)) +
  ggpubr::stat_cor(method = "spearman", label.x = 0) +
  theme_classic() +
  xlab("Signature Score 1") +
  ylab("Signature Score 2") +
  theme(axis.title = element_text(size = 14),
        legend.position = "none") +
  ggtitle(paste0("Correlation between signature scores across cell types")
          ) -> p0
ggplotly(p0)
```

### Correlation between signature scores across cell types

```{r message=FALSE, warning=FALSE}
d1 %>%
  dplyr::filter(!grepl("Patient", cell_type)) %>%
  ggplot(aes(x = gene1, y = gene2)) +
  geom_point(size = 1, aes(color = cell_type)) +
  stat_smooth(method = "lm", se = TRUE, fill = "gray", color = "darkgray",
              formula = y ~ poly(x, 1, raw = TRUE)) +
  ggpubr::stat_cor(method = "spearman", label.x = 0) +
  theme_classic() +
  xlab("Signature Score 1") +
  ylab("Signature Score 2") +
  theme(axis.title = element_text(size = 14),
        legend.position = "none") +
  ggtitle(paste0("Correlation between signature scores across cell types")) +
  facet_grid(cols = vars(cell_type)) -> p4
# p4
ggplotly(p4)
```

```{r}
sessioninfo::session_info()
```
